---
title: "Omus_sympatry_visualization"
author: "Maddy Enda"
format: html
editor: visual
---

# Creating a visualization for the Omus sympatry distribution report

## Author: Maddy Enda

## Contact: madisonenda\@gmail.com

## Last Edit: 06/05/2026

### Overview:

-   In this script, we will be creating a geospatial visualization for a report on the distribution of `Omus sympatry`.

### Import libraries

```{r}
# Import relevant libraries
librarian::shelf(tidyverse,
                 dplyr,
                 stars,
                 sf,
                 readxl,
                 scatterpie,
                 ggtext,
                 patchwork,
                 ggspatial,
                 cowplot,
                 ggpattern,
                 ggimgae,
                 calecopal)
```

### Import the data

```{r}
# Load in the excel file
Omus_forms <- read_excel("C:/Users/Madison.Enda/Desktop/omus_data/Omus_forms.xlsx")
```

### Replace NA with 0

```{r}
# Replace NAs
Omus_forms[is.na(Omus_forms)] <- 0

# Create a species percentage column
Omus_forms <- Omus_forms %>%
  mutate(
    lec_pct = lecontei / Total,
    cal_pct = californicus / Total,
    cyl_pct = subcylindricus / Total,
    met_pct = submetallicus / Total
  )
```

```{r}
### Add points
# Create points from your coordinate dataframe
Omus_forms <- Omus_forms %>%
  separate(
    Coordinates,
    into = c("lat", "lon"),
    sep = ","
  ) %>%
  mutate(
    lat = as.numeric(trimws(lat)),
    lon = as.numeric(trimws(lon))
  )

# Convert the lat and lon columns to points
pts <- st_as_sf(
  Omus_forms,
  coords = c("lon", "lat"),
  crs = 4326
)
```

### Read in the california shapefile

```{r}
# Read in shapefile
shp <- st_read("C:/Users/Madison.Enda/Desktop/omus_data/CA_Counties.shp")

# Convert points to be same
pts <- st_transform(pts, st_crs(shp))
```

### Create bounding box

```{r}
bb <- st_bbox(pts)

bb_poly <- st_as_sfc(bb, crs = st_crs(pts))

bb_poly_proj <- st_transform(bb_poly, st_crs(shp))

shp_clip <- st_crop(shp, st_bbox(bb_poly_proj))
```

### Make pie charts

```{r}
# ----------------------------
# 1. Aggregate by county
# ----------------------------
county_summary <- Omus_forms %>%
  group_by(County) %>%
  summarise(
    lecontei = sum(lecontei, na.rm = TRUE),
    subcylindricus = sum(subcylindricus, na.rm = TRUE),
    submetallicus = sum(submetallicus, na.rm = TRUE),
    californicus = sum(californicus, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rowwise() %>%
  mutate(
    Total = sum(c_across(c(lecontei, subcylindricus, submetallicus, californicus))),
    lecontei = lecontei / Total,
    subcylindricus = subcylindricus / Total,
    submetallicus = submetallicus / Total,
    californicus = californicus / Total
  ) %>%
  ungroup()

# ----------------------------
# 2. Long format
# ----------------------------
county_long <- county_summary %>%
  pivot_longer(
    cols = c(lecontei, subcylindricus, submetallicus, californicus),
    names_to = "species",
    values_to = "pct"
  )

county_long$species <- as.character(county_long$species)

county_long$species <- factor(
  county_long$species,
  levels = c("lecontei", "subcylindricus", "submetallicus", "californicus")
)

# ----------------------------
# 3. Pattern mapping
# ----------------------------
pattern_types <- c(
  lecontei = "stripe",
  subcylindricus = "crosshatch",
  submetallicus = "circle",
  californicus = "none"
)
```

### Built pie charts

```{r}
# Creating custom colors
species_colors <- c("#66CC66", "lightblue", "#E6D2B5", "#F7BB38")

names(species_colors) <- c(
  "lecontei",
  "subcylindricus",
  "submetallicus",
  "californicus"
)
```

### Use new font

```{r}
install.packages(c("showtext", "sysfonts"))
library(showtext)
library(sysfonts)

font_add_google("Montserrat", "montserrat")
font_add_google("Inter", "inter")
font_add_google("Open Sans", "opensans")
font_add_google("Public Sans", "publicsans")

showtext_auto()
```

### Make pie chart labels

```{r}
county_long <- county_long %>%
  mutate(label = paste0(round(pct * 100, 1), "%"))

county_long <- county_long %>%
  group_by(County) %>%
  mutate(ypos = cumsum(pct) - 0.5 * pct) %>%
  ungroup()
```

```{r}
# ----------------------------
# 4. Pie function (PATTERN ONLY, NO COLOR)
# ----------------------------
make_pie <- function(df){

  ggplot(df, aes(x = "", y = pct, fill = species, pattern = species)) +
    
   geom_col_pattern(
  color = "black",
  width = 1,
  linewidth = 0.18,
  linejoin = "round",

  pattern_fill = "#FDF5E6",
  pattern_colour = "#FDF5E6",
  pattern_density = 0.1,
  pattern_spacing = 0.03
) +
    
    geom_text(
      data = ~ dplyr::filter(.x, pct > 0),
      aes(label = paste0(round(pct * 100, 1), "%")),
      position = position_stack(vjust = 0.5),
      color = "black",
      fontface = "bold",
      family = "inter",
      size = 8.0
    ) +
    
    coord_polar(theta = "y", direction = -1, clip = "on") +
    
    scale_fill_manual(values = species_colors) +
    scale_pattern_manual(values = pattern_types) +
    
    theme_void() +
    theme(
      plot.title = element_text(
        size = 28,
        family = "montserrat",
        face = "bold"
      ),
      plot.margin = margin(0, 0, 0, 0),
      legend.position = "none",
      plot.background = element_rect(fill = "transparent", color = NA),
  panel.background = element_rect(fill = "transparent", color = NA)
    ) +
    
    ggtitle(unique(df$County))
}
# ----------------------------
# 5. Build pies
# ----------------------------
pie_list <- split(county_long, county_long$County)
pies <- lapply(pie_list, make_pie)

pie_panel <- wrap_plots(pies, ncol = 2)
```

```{r}
pie_panel
```

### save it

```{r}
pie_panel <- pie_panel &
  theme(
    plot.background = element_rect(fill = "transparent", color = NA)
  )

ggsave(
  "pies.png",
  pie_panel,
  bg = "transparent",
  width = 12,
  height = 8,
  dpi = 300
)
```

### Make title object

```{r}
title_plot <- ggplot() +
  annotate(
    "text",
    x = 0.5,
    y = 0.5,
    label = "Distribution of Omus Species Counts Across Central California Counties",
    family = "inter",
    fontface = "bold",
    size = 20
  ) +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA)
  )
```

```{r}
title_plot
```

### Save Title

```{r}
ggsave(
  "title.png",
  title_plot,
  bg = "transparent",
  width = 10,
  height = 3,
  dpi = 300
)
```

### Make a legend

```{r}
species_vec <- names(species_colors)
n <- length(species_vec)


legend_df <- data.frame(
  species = species_vec,
  x = seq(1, by = 0.6, length.out = n),
  y = 1
)

legend_panel <- ggplot(
  legend_df,
  aes(
    x = x,
    y = y,
    fill = species,
    pattern = species
  )
) +
  geom_tile_pattern(
    width = 0.25,
    height = 0.25,
    color = "black",
    pattern_fill = "#FDF5E6",
    pattern_colour = "#FDF5E6",
    pattern_density = 0.4,
    pattern_spacing = 0.03
  ) +

  # species names just to the right of each box
  geom_text(
  aes(
    x = x,
    y = 0.80,
    label = species
  ),
  hjust = 0.5,
  vjust = 1,
  family = "inter",
  fontface = "italic",
  size = 8
) +

  # title
  annotate(
    "text",
    x = mean(range(legend_df$x)),
    y = 1.3,
    label = "Omus species",
    family = "inter",
    fontface = "bold",
    size = 15
  ) +

  scale_fill_manual(values = species_colors) +
  scale_pattern_manual(values = pattern_types) +

  coord_fixed(
  clip = "off",
  xlim = c(0.5, length(species_colors) + 1.5),
  ylim = c(0.4, 2)   # key fix for title + labels
) + 

  theme_void() +
 theme(
  plot.background = element_rect(fill = "transparent", color = NA),
  panel.background = element_rect(fill = "transparent", color = NA),
  legend.background = element_rect(fill = "transparent", color = NA),
  legend.position= "none",
  panel.grid = element_blank()
)
```

```{r}
plot(legend_panel)
```

### Save legend

```{r}
showtext_auto(TRUE)

ggsave(
  "legend.png",
  legend_panel,
  bg = "transparent",
  width = 10,
  height = 4,
  dpi = 300
)
```

### Load in beetle png

```{r}
# Load in little beetle
beetle_png <- "ground-beetle.png"

# Add image path for pts
pts$img <- beetle_png

# plot little beetle
coords <- st_coordinates(pts)

pts_img <- cbind(pts, coords)
```

### Create main map

```{r}
# Highlight our counties
target_counties <- c(
  "Fresno",
  "Merced",
  "Monterey",
  "San Benito",
  "San Francisco",
  "San Luis Obispo",
  "San Mateo",
  "Santa Barbara",
  "Santa Clara",
  "Santa Cruz"
)

green_outline <- shp_clip %>%
  dplyr::filter(NAME %in% target_counties)

# ----------------------------
# MAIN MAP
# ----------------------------
central_map <- ggplot() +
  geom_sf(
    data = shp_clip,
    fill = "grey95",
    color = "grey50",
    linewidth = 0.3
  ) +
  geom_sf(
    data = green_outline,
    fill = NA,
    color = "black",
    linewidth = 1.0,
  ) +
  geom_image(
    data = pts_img,
    aes(x = X, y = Y, image = img),
    size = 0.06
  ) +
  coord_sf(clip = "off") +
  theme_void() +
  theme(
    plot.margin = margin(20, 20, 20, 20)
  )

# ----------------------------
# CALIFORNIA INSET
# ----------------------------
locator_map <- ggplot() +
  geom_sf(
    data = shp,
    fill = "grey70",
    color = "grey40"
  ) +
  theme_void()

bbox <- st_bbox(shp_clip)

locator_map <- locator_map +
  annotate(
    "rect",
    xmin = bbox["xmin"],
    xmax = bbox["xmax"],
    ymin = bbox["ymin"],
    ymax = bbox["ymax"],
    fill = NA,
    color = "green2",
    linewidth = 1
  )

map_panel <-
  central_map +
  inset_element(
    locator_map,
    left = 0.57,
    bottom = 0.70,
    right = 0.95,
    top = 0.95
  )
```

```{r}
map_panel <- map_panel &
  theme(
    plot.background = element_rect(fill = "transparent", color = NA)
  )

plot(map_panel)
```

### Save the map

```{r}
# Publication quality
ggsave(
  "omus_distribution_map_2.png",
  map_panel,
  width = 7,
  height = 5,
  units = "in",
  dpi = 600
)
```

# Arrange final product!

### Load libraries

```{r}
library(png)
library(grid)
```

```{r}
# create base
base <- ggplot() +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "#F6F1E6", color = NA)
  )
```

```{r}
# Read in pngs
title_img <- rasterGrob(readPNG("title.png"), interpolate = TRUE)
map_img   <- rasterGrob(readPNG("omus_distribution_map_2.png"), interpolate = TRUE)
pies_img  <- rasterGrob(readPNG("pies.png"), interpolate = TRUE)
legend_img <- rasterGrob(readPNG("legend.png"), interpolate = TRUE)
```

```{r}
final <- base +

  # -----------------------
  # TITLE (top 10%)
  # -----------------------
  annotation_custom(
    title_img,
    xmin = 0.05, xmax = 0.95,
    ymin = 0.90, ymax = 0.99
  ) +

  # -----------------------
  # MAP (60% width, full height under title)
  # -----------------------
  annotation_custom(
    map_img,
    xmin = 0.00, xmax = 0.60,
    ymin = 0.00, ymax = 0.90
  ) +

  # -----------------------
  # PIES (top-right block)
  # -----------------------
  annotation_custom(
    pies_img,
    xmin = 0.60, xmax = 1.00,
    ymin = 0.40, ymax = 0.90
  ) +

  # -----------------------
  # LEGEND (bottom-right under pies)
  # -----------------------
  annotation_custom(
    legend_img,
    xmin = 0.60, xmax = 1.00,
    ymin = 0.00, ymax = 0.40
  )
```

```{r}
final
```
