---
title: "Omus_sympatry_visualization"
author: "Maddy Enda"
format: html
editor: visual
---

# Creating a visualization for the Omus sympatry distribution report

## Author: Maddy Enda

## Contact: madisonenda\@gmail.com

## Last Edit: 06/05/2026

### Overview:

-   In this script, we will be creating a geospatial visualization for a report on the distribution of `Omus sympatry`.

### Import libraries

```{r}
# Import relevant libraries
librarian::shelf(tidyverse,
                 dplyr,
                 stars,
                 sf,
                 readxl,
                 scatterpie,
                 ggtext,
                 patchwork,
                 ggspatial,
                 cowplot,
                 ggpattern,
                 ggimgae,
                 calecopal)
```

### Import the data

```{r}
# Load in the excel file
Omus_forms <- read_excel("C:/Users/Madison.Enda/Desktop/omus_data/Omus_forms.xlsx")
```

### Replace NA with 0

```{r}
# Replace NAs
Omus_forms[is.na(Omus_forms)] <- 0

# Create a species percentage column
Omus_forms <- Omus_forms %>%
  mutate(
    lec_pct = lecontei / Total,
    cal_pct = californicus / Total,
    cyl_pct = subcylindricus / Total,
    met_pct = submetallicus / Total
  )
```

```{r}
### Add points
# Create points from your coordinate dataframe
Omus_forms <- Omus_forms %>%
  separate(
    Coordinates,
    into = c("lat", "lon"),
    sep = ","
  ) %>%
  mutate(
    lat = as.numeric(trimws(lat)),
    lon = as.numeric(trimws(lon))
  )

# Convert the lat and lon columns to points
pts <- st_as_sf(
  Omus_forms,
  coords = c("lon", "lat"),
  crs = 4326
)
```

### Read in the california shapefile

```{r}
# Read in shapefile
shp <- st_read("C:/Users/Madison.Enda/Desktop/omus_data/CA_Counties.shp")

# Convert points to be same
pts <- st_transform(pts, st_crs(shp))
```

### Create bounding box

```{r}
bb <- st_bbox(pts)

bb_poly <- st_as_sfc(bb, crs = st_crs(pts))

bb_poly_proj <- st_transform(bb_poly, st_crs(shp))

shp_clip <- st_crop(shp, st_bbox(bb_poly_proj))
```

### Make pie charts

```{r}
# ----------------------------
# 1. Aggregate by county
# ----------------------------
county_summary <- Omus_forms %>%
  group_by(County) %>%
  summarise(
    lecontei = sum(lecontei, na.rm = TRUE),
    subcylindricus = sum(subcylindricus, na.rm = TRUE),
    submetallicus = sum(submetallicus, na.rm = TRUE),
    californicus = sum(californicus, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rowwise() %>%
  mutate(
    Total = sum(c_across(c(lecontei, subcylindricus, submetallicus, californicus))),
    lecontei = lecontei / Total,
    subcylindricus = subcylindricus / Total,
    submetallicus = submetallicus / Total,
    californicus = californicus / Total
  ) %>%
  ungroup()

# ----------------------------
# 2. Long format
# ----------------------------
county_long <- county_summary %>%
  pivot_longer(
    cols = c(lecontei, subcylindricus, submetallicus, californicus),
    names_to = "species",
    values_to = "pct"
  )

county_long$species <- as.character(county_long$species)

county_long$species <- factor(
  county_long$species,
  levels = c("lecontei", "subcylindricus", "submetallicus", "californicus")
)

# ----------------------------
# 3. Pattern mapping
# ----------------------------
pattern_types <- c(
  lecontei = "stripe",
  subcylindricus = "crosshatch",
  submetallicus = "circle",
  californicus = "none"
)
```

### Built pie charts

```{r}
# Creating custom colors
species_colors <- c("#66CC66", "lightblue", "#E6D2B5", "#F7BB38")

names(species_colors) <- c(
  "lecontei",
  "subcylindricus",
  "submetallicus",
  "californicus"
)
```

### Use new font

```{r}
install.packages(c("showtext", "sysfonts"))
library(showtext)
library(sysfonts)

font_add_google("Montserrat", "montserrat")
font_add_google("Inter", "inter")
font_add_google("Open Sans", "opensans")
font_add_google("Public Sans", "publicsans")

showtext_auto()
```

### Make pie chart labels

```{r}
county_long <- county_long %>%
  mutate(label = paste0(round(pct * 100, 1), "%"))

county_long <- county_long %>%
  group_by(County) %>%
  mutate(ypos = cumsum(pct) - 0.5 * pct) %>%
  ungroup()
```


```{r}
# ----------------------------
# 4. Pie function (PATTERN ONLY, NO COLOR)
# ----------------------------
make_pie <- function(df){

  ggplot(df, aes(x = "", y = pct, fill = species, pattern = species)) +
    
    geom_col_pattern(
      color = "black",
      width = 1,
      
      # keep your colors
      pattern_fill = "#FDF5E6",
      pattern_colour = "#FDF5E6",
      pattern_density = 0.4,
      pattern_spacing = 0.03
    ) +
    
    geom_text(
      data = ~ dplyr::filter(.x, pct > 0),
      aes(label = paste0(round(pct * 100, 1), "%")),
      position = position_stack(vjust = 0.5),
      color = "black",
      fontface = "bold",
      family = "inter",
      size = 3.0
    ) +
    
    coord_polar(theta = "y", direction = -1) +
    
    scale_fill_manual(values = species_colors) +
    scale_pattern_manual(values = pattern_types) +
    
    theme_void() +
    theme(
      plot.title = element_text(
        size = 8,
        family = "montserrat",
        face = "bold"
      ),
      plot.margin = margin(0, 0, 0, 0),
      legend.position = "none"
    ) +
    
    ggtitle(unique(df$County))
}
# ----------------------------
# 5. Build pies
# ----------------------------
pie_list <- split(county_long, county_long$County)
pies <- lapply(pie_list, make_pie)

pie_panel <- wrap_plots(pies, ncol = 2)
```

```{r}
pie_panel
```
### save it

```{r}
pie_panel <- pie_panel &
  theme(
    plot.background = element_rect(fill = "transparent", color = NA)
  )

ggsave(
  "pie_charts.png",
  plot = pie_panel,
  width = 7,
  height = 5,
  dpi = 600,
  bg = "transparent"
)
```


### Make a legend

```{r}
# ----------------------------
# 6. Create ONE clean legend (pattern only)
# ----------------------------
legend_df <- data.frame(
  species = levels(county_long$species),
  pct = 1
)

legend_plot <- ggplot(
  legend_df,
  aes(x = species, y = pct, pattern = species)
) +
  geom_col_pattern(
    fill = "white",        # IMPORTANT
    color = "black",
    pattern_fill = "black",
    pattern_colour = "black"
  ) +
  scale_pattern_manual(values = pattern_types) +
  coord_flip() +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9)
  ) +
  guides(
    fill = "none",
    pattern = guide_legend(title = "Species")
  )

# Extract shared legend
shared_legend <- get_legend(legend_plot)

```


### Load in beetle png

```{r}
# Load in little beetle
beetle_png <- "ground-beetle.png"

# Add image path for pts
pts$img <- beetle_png

# plot little beetle
coords <- st_coordinates(pts)

pts_img <- cbind(pts, coords)
```


### Create main map

```{r}
# Highlight our counties
target_counties <- c(
  "Fresno",
  "Merced",
  "Monterey",
  "San Benito",
  "San Francisco",
  "San Luis Obispo",
  "San Mateo",
  "Santa Barbara",
  "Santa Clara",
  "Santa Cruz"
)

green_outline <- shp_clip %>%
  dplyr::filter(NAME %in% target_counties)

# ----------------------------
# MAIN MAP
# ----------------------------
central_map <- ggplot() +
  geom_sf(
    data = shp_clip,
    fill = "oldlace",
    color = "grey50",
    linewidth = 0.3
  ) +
  geom_sf(
    data = green_outline,
    fill = NA,
    color = "green4",
    linewidth = 1.0,
  ) +
  geom_image(
    data = pts_img,
    aes(x = X, y = Y, image = img),
    size = 0.06
  ) +
  coord_sf(clip = "off") +
  theme_void() +
  theme(
    plot.margin = margin(20, 20, 20, 20)
  )

# ----------------------------
# CALIFORNIA INSET
# ----------------------------
locator_map <- ggplot() +
  geom_sf(
    data = shp,
    fill = "burlywood3",
    color = "grey40"
  ) +
  theme_void()

bbox <- st_bbox(shp_clip)

locator_map <- locator_map +
  annotate(
    "rect",
    xmin = bbox["xmin"],
    xmax = bbox["xmax"],
    ymin = bbox["ymin"],
    ymax = bbox["ymax"],
    fill = NA,
    color = "green2",
    linewidth = 1
  )

map_panel <-
  central_map +
  inset_element(
    locator_map,
    left = 0.57,
    bottom = 0.70,
    right = 0.95,
    top = 0.95
  )
```

```{r}
map_panel <- map_panel &
  theme(
    plot.background = element_rect(fill = "transparent", color = NA)
  )

plot(map_panel)
```
### Save the map

```{r}
# Publication quality
ggsave(
  "omus_distribution_map_2.png",
  map_panel,
  width = 7,
  height = 5,
  units = "in",
  dpi = 600
)
```


