---
title: "Omus_sympatry_visualization"
author: "Maddy Enda"
format: html
editor: visual
---

# Creating a visualization for the Omus sympatry distribution report

## Author: Maddy Enda
## Contact: madisonenda@gmail.com
## Last Edit: 06/05/2026


### Overview:

- In this script, we will be creating a geospatial visualization for a report on the distribution of `Omus sympatry`.


### Import libraries

```{r}
# Import relevant libraries
librarian::shelf(tidyverse,
                 dplyr,
                 stars,
                 sf,
                 readxl,
                 tmap)
```

### Import the data

```{r}
# Load in the excel file
Omus_forms <- read_excel("C:/Users/Madison.Enda/Desktop/omus_data/Omus_forms.xlsx")
```

### Replace NA with 0

```{r}
# Replace NAs
Omus_forms[is.na(Omus_forms)] <- 0
```

### Read in the california shapefile

```{r}
# Read in shapefile
shp <- st_read("C:/Users/Madison.Enda/Desktop/omus_data/CA_Counties.shp")
```

### Quick visual

```{r}
# Visualize
tm_shape(shp) +
  tm_polygons()
```
### Add points

```{r}
# Create points from your coordinate dataframe
Omus_forms <- Omus_forms %>%
  separate(
    Coordinates,
    into = c("lat", "lon"),
    sep = ","
  ) %>%
  mutate(
    lat = as.numeric(trimws(lat)),
    lon = as.numeric(trimws(lon))
  )

# Convert the lat and lon columns to points
pts <- st_as_sf(
  Omus_forms,
  coords = c("lon", "lat"),
  crs = 4326
)

pts <- st_transform(pts, st_crs(shp))
```

### Create bounding box

```{r}
bb <- st_bbox(pts)

bb_poly <- st_as_sfc(bb, crs = st_crs(pts))

bb_poly_proj <- st_transform(bb_poly, st_crs(shp))

shp_clip <- st_crop(shp, st_bbox(bb_poly_proj))
```

### Test new clip

```{r}
tmap_mode("plot")

tm_shape(shp_clip) +
  tm_polygons(fill = "gray90") +
  tm_shape(pts) +
  tm_dots(
    fill = "SiteID",
    palette = "Set3",
    size = 0.15
  )
```
```{r}
tm_shape(shp_clip) +
  tm_polygons(fill = "gray90") +
tm_shape(pts) +
  tm_dots(size = 0.5, fill = "Site")
```




